Converging High-Throughput and High-Performance Computing Paradigms: \\ A Case Study of Using Abstractions to Bridge the Divide

### 2017-03-17 PandaEvolution.tex
Iteration and shortening IV and V; add new data from Sergey

The next LHC run for taking data (Run 3) will require more resources than the
Worldwide LHC Computing Grid (WLCG) can possibly provide. Currently, PanDA WMS
uses more than 100,000 cores at over 100 Grid sites, with a peak performance of
0.3 petaFLOPS. This capacity will be sufficient for the planned analysis and
data processing, but it will be insufficient for the Monte Carlo production
workflow and any extra activity. To alleviate these challenges, ATLAS is engaged
in a program to expand the current computing model to include additional
resources such as the opportunistic use of supercomputers as well as commercial
and academic clouds.

@@

Modern supercomputers have been designed mainly to support parallel computation
that requires runtime communication. Job execution is parallelized across
multiple cores, each core calculating a small part of the problem and
communicating with other cores via a message passing interface (MPI).
Accordingly, supercomputers have large number of worker nodes, connected through
a high-speed, low-latency dedicated network. Each worker node has multicore
CPUs, usually augmented with massively parallel Graphics Processing Units (GPUs)
or other types of specialized coprocessors.

PanDA WMS has been designed to support distributed Grid
computing\sergeynote{specifically "distributed Grid computing"}\mtnote{done.}.
Executing ATLAS workloads or workflows involves concurrent and/or sequential
runs of possibly large amount of jobs, each requiring no or minimal
parallelization and no runtime communication. Thus, computing infrastructure
like WLCG have been designed to aggregate large amount of computing resources
across multiple sites.
%, often geographically dislocated.
\sergeynote{geographically distributed?}\mtnote{Better?} While each site may
deploy runtime message-passing capabilities, usually these are not used to
perform distributed computations.

@@

The Titan supercomputer, current number three
% (number one from November 2012 until June 2013)
on the Top 500 list~\cite{top500}, is a Cray XK7 system with 18,688 work nodes
and a total of 299,008 CPU cores. Each work node has an AMD Opteron CPU, a
Nvidia Tesla GPU, 32 GB of RAM and no local storage, though a 16 GB RAM disk can
be set up. Work nodes use Cray’s Gemini interconnect for inter-node MPI
messaging. Titan is served by the Spider II~\cite{spider2}, a Lustre filesystem
with 32 PB of disk storage, and by a 29 PB HPSS tape storage system. Titan’s
work nodes run Compute Node Linux, a run time environment based on SUSE Linux
Enterprise Server.

% Titan, was the first large-scale system to use a hybrid architecture that
% utilizes worker nodes with both AMD 16-core Opteron 6274 CPUs and NVIDIA Tesla
% K20 GPU accelerators.
% This hybrid design provides improved energy efficiency, as well as an order of
% magnitude in computational capacity over its predecessor.

Titan offers to its users login nodes and data transfer nodes (DTNs). Users have
to log into one of these nodes via ssh to submit jobs to Titan's PBS scheduler.
Titan's authentication and authorization model is based on two-factor
authentication with a RSA SecurID key, generated every 30 seconds. Login nodes
and DTNs are the only nodes with out/inbound wide area network connectivity.
Work nodes have only local network access. Data staging is supported via the
DTNs, from the wide area network and the OLCF high performance storage system
(HPSS). DTNs are shared across all Titan's users that can schedule jobs to a
dedicates batch system to automate data staging. Fair-share policies are in
place both for the batch system and for the use of DTN resources.

Titan's architecture, configuration and policies poses several challenges to the
integration with PanDA. PanDA Pilot requires to contact the Job Dispatcher
module of the PanDA Server to pull jobs to execute. Titan's work nodes do not
offer outbound network connectivity, making the current deployment model of
PanDA pilots unfeasible. Titan does not support PanDA's security model based on
certificates and virtual organizations, making the PanDA's approach to identity
management also unfeasible. While DTNs offer wide area network data transfer, an
integration with ATLAS DDM is beyond the functional and administrative scope of
the current prototyping phase. Finally, the specific characteristics of the
execution environment, especially the absence of local storage on the work nodes
and modules tailored to Compute Node Linux, require reengineering of ATLAS
application frameworks.

Currently, no HEP application can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks can be offloaded from
the Grid to Titan's large amount of cores. Further, when HEP tasks can be
partitioned to independent jobs, Titan work nodes can be used to execute up to
16 concurrent jobs' payload, one for each available core. Given these
constraints and challenges, the type of task most suitable for execution at the moment on Titan is Monte Carlo
% event generation and
detector simulation.
% Both types of task are
This type of task is mostly computational-intensive, requiring less than 2GB of
RAM at runtime and with small input data requirements. Detector simulation tasks
in ATLAS account for XX--XX\% of all jobs on WLCG, corresponding to X/X of the
current Grid capacity.\mtnote{Please provide numbers to replace `XX' and,
possibly, a diagram illustrating detector simulation stats compared to the whole
ATLAS computation activity.}

% The event generation in ATLAS accounts for 10--15\% of all jobs on WLCG,
% corresponding to 2/3 of the current Grid capacity.

% Event generation and
Detector simulation is part of the ATLAS production Monte Carlo (MC) workflow
(also known as MC production
chain)~\cite{rimoldi2006atlas,de2013delphes,ritsch2014atlas}. The MC workflow
consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Even generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{sjostrand2006pythia}
and HERWIG~\cite{corcella2001herwig}. The detector simulator is called
Geant4~\cite{agostinelli2003geant4} and simulates the interaction of these
particles with the sensitive material of the ATLAS detector. Each interaction
creates a so-called hit and all hits are collected and passed on for
digitalization where hits are further process to mimic the readout of the
detector. Finally, reconstruction operates local pattern recognition, creating
high-level objects like particles and jets.

@@

The lack of wide area network connectivity on Titan's work nodes is the most
relevant challenge for integrating PanDA WMS and Titan. Without connectivity,
Panda Pilots cannot be scheduled on work nodes because they would not be able to
communicate with PanDA Server and therefore pull and execute jobs. This makes
impossible to port PanDA Pilot to Titan while maintaining the defining feature
of the pilot abstraction: decoupling resource acquisition from workload
execution via multi-stage scheduling.

The unavailability of pilots is a potential drawback when executing distributed
workloads like MC
% event generation and
detector simulation. Pilots are used to increase the throughput of distributed
workloads: while pilots have to wait in the supercomputer's queue, once
scheduled, they can pull and execute jobs independently from the system's queue.
Jobs can be concurrently on every core available to the pilot and multiple
generations of concurrent executions can be performed until the pilot's walltime
is exhausted. This is particularly relevant for machines like Titan where queue
policies privilege parallel jobs on the base of the number of work nodes they
request: the higher the number of nodes, the shorter the amount of queue time
(modulo fair share policies).

@@

Titan’s backfill functionality offers the opportunity to avoid the overhead of
queue wait times without using pilot abstraction. Backfill availability is the
number of work nodes that cannot be used by any of the jobs already queued on
Titan: All queued jobs are either too large or their walltime is too long. At
any point in time, Titan’s Moab scheduler can be queried for backfill
availability. Based upon the result of this query, a job can be shaped to
request no more than the backfill availability, and when submitted as such is
usually scheduled immediately spending almost no time in the queue.

Compared to pilots, backfill has the disadvantage of limiting the amount of work
nodes that can be requested. Pilots are normal jobs: they can request as many
work nodes for as much time as a queue can offer. On the contrary, jobs sized on
the basis of backfill information availability \jhanote{should be resource
availability or backfill queue information?}\mtnote{AFAIH, there is no dedicated
queue to backfill} depend on the number % amount
of work nodes that cannot be given to any other job. any other job in the
Titan's queue\sergeynote{any other job currently in the Titan's
queue?}\mtnote{Better?}. Usually, backfill availability is a fraction of the
total capacity of the queue but the size of Titan mitigates this limitation.
Every year, about 10\% of Titan's capacity remains
unused~\cite{titan_utilization}, corresponding to an average of 30,000 unused
cores. This equals approximately 270M core hours per year, roughly 30\% of the
overall capacity of WLCG.

% A system that can use those temporarily free work nodes with smaller
% scale tasks would be very valuable. For this reason, Titan offers backfill
% functionalities: users can interrogate Titan's Moab scheduler about how many
% free nodes are currently available and for how long. If a user submit a
% requestvfor that number of nodes and walltime, queue time will be minimal.

@@

Given the communication requirements of PanDA Pilots and the unused capacity of
Titan, PanDA pilot was redesigned to serve as a job broker on Titan. Architected
as a stand-alone service to be executed on a DTN, this prototype called `PanDA
Broker' implements functionalities to: (i) interrogate Titan about backfill
availability; (ii) pull MC jobs and events; (iii) wrap jobs' payload into PBS
MPI jobs; (iv) submitting MPI PBS jobs to Titan's PBS batch system and monitor
their execution;\sergeynote{Monitor job state and state transitions using
SAGA}\mtnote{Better?} and (v) staging input/output files. Backfill querying,
payload wrapping, and job submission required a new implementation while job and
events pulling, and file staging were mostly inherited from PanDA Pilot.

@@

Backfill querying is performed via a dedicated Moab command while a tailored
Python MPI script is used to wrap job's payload. This MPI wrapper serves as a
``container'' for non-MPI jobs, enabling execution of unmodified Grid-centric
jobs on Titan. Typically, a MPI wrapper is workload-specific as it sets up the
execution environment for a specific payload. This involves organization of
worker directories, data management, optional input parameters modification, and
cleanup on exit. Upon submission, a copy of the MPI wrapper runs on every
available work node. Each script knows its MPI rank (an index that runs from
zero to a maximum number of nodes or script copies) as well as the total number
of ranks in the current submission. When activated on worker nodes, each copy of
the wrapper script starts the execution of the payload in a subprocess and waits
until its completion.

MPI wrappers are submitted to Titan's PBS batch system via
RADICAL-SAGA~\cite{radical-saga}, a Python module, compliant with the OGF GFD.90
SAGA specification~\cite{saga-spec}. The Simple API for Grid Applications (SAGA)
offers a unified interface to diverse job schedulers and file transferring
services. In this way, SAGA provides an interoperability layer that lowers the
complexity of using distributed infrastructures. Behind the API façade,
RADICAL-SAGA implements a adaptor architecture: each adaptor interface the SAGA
API with different middleware systems and services, including the batch
scheduler of Titan.

The data staging capabilities of the PanDA Broker are implemented via a file
system that is shared among DTNs and work nodes. The input files required by the
payload are downloaded via the ATLAS DDM service and then stored in a directory
on the shared filesystem. The MPI wrapper setup process includes making the
location of these files available to the payload. The PanDA Broker can locate
the Payloads' output files on the shared filesystem and transfer them from Titan
to the ATLAS Tier 1 computing center at Brookhaven National Lab (BNL) or any
other Grid site.

Once deployed, every PanDA Broker supports the execution of MC detector
simulation on Titan in 8 steps: (i) PanDA Broker queries Titan's Moab scheduler
about current backfill availability; (ii) the PanDA Broker queries the Job
Dispatcher module of the PanDA server for jobs that have been bound to Titan by
JEDI; (iii)  the PanDA Broker requests to the Data Service module of the PanDA
Server what input data these jobs require and transfer the files to the DTN via
the ATLAS DDM service; (iv) the PanDA Broker creates a PBS MPI script, wrapping
enough jobs' payload to fit backfill availability; (v) the PanDA Broker submits
the PBS MPI script to the Titan's PBS batch system via RADICAL-SAGA; (vi) upon
execution on the worker node(s), the MPI script creates a ramdisk and configure
AthenaMP to execute the wrapped payload; (vii) the payload is executed into
dedicated processes, 16 for each work node, 1 for each available core; (viii)
upon completion of the MPI script, the PanDA Broker locates the output on the
shared NFS filesystem and transfer it to BNL ATLAS Tier 1 computing centre.

>>>>

The next LHC run for taking data (Run 3) will require more resources than the
Worldwide LHC Computing Grid (WLCG) can provide. Currently, PanDA WMS uses more
than 100,000 cores at over 100 Grid sites, with a peak performance of 0.3
petaFLOPS. This capacity will be sufficient for the planned analysis and data
processing, but it will be insufficient for the Monte Carlo production workflow
and any extra activity. To alleviate these challenges, ATLAS is engaged in a
program to expand the current computing model to include additional resources
such as the opportunistic use of supercomputers as well as commercial and
academic clouds.

@@

Modern supercomputers have been designed mainly to support parallel computation
that requires runtime communication. Job execution is parallelized across
multiple cores, each core calculating a small part of the problem and
communicating with other cores via a message passing interface (MPI).
Accordingly, supercomputers have large number of worker nodes, connected through
a high-speed, low-latency dedicated network. Each worker node has multicore
CPUs, usually augmented with parallel Graphics Processing Units (GPUs) or other
types of specialized coprocessors.

% \sergeynote{specifically "distributed Grid computing"}\mtnote{done.}.
% \sergeynote{geographically distributed?}\mtnote{Better?}

PanDA WMS has been designed to support distributed Grid computing. Executing
ATLAS workloads or workflows involves concurrent and/or sequential runs of
possibly large amount of jobs, each requiring no or minimal parallelization and
no runtime communication. Thus, computing infrastructure like WLCG have been
designed to aggregate large amount of computing resources across multiple sites.
While each site may deploy runtime message-passing capabilities, usually these
are not used to perform distributed computations.

@@

The Titan supercomputer, current number three on the Top 500 list~\cite{top500},
is a Cray XK7 system with 18,688 worker nodes and a total of 299,008 CPU cores.
Each worker node has an AMD Opteron  6274 16-core CPU, a Nvidia Tesla K20X GPU,
32 GB of RAM and no local storage, though a 16 GB RAM disk can be set up. Work
nodes use Cray’s Gemini interconnect for inter-node MPI messaging. Titan is
served by the Spider II~\cite{spider2}, a Lustre filesystem with 32 PB of disk
storage, and by a 29 PB HPSS tape storage system. Titan’s worker nodes run
Compute Node Linux, a run time environment based on SUSE Linux Enterprise
Server.

% Titan, was the first large-scale system to use a hybrid architecture that
% utilizes worker nodes with both AMD 16-core Opteron 6274 CPUs and NVIDIA Tesla
% K20 GPU accelerators. This hybrid design provides improved energy efficiency,
% as well as an order of magnitude in computational capacity over its
% predecessor.

Titan's users submit jobs to Titan's PBS scheduler by logging into login or data
transfer nodes (DTNs). Titan's authentication and authorization model is based
on two-factor authentication with a RSA SecurID key, generated every 30 seconds.
Login nodes and DTNs have out/inbound wide area network connectivity while
worker nodes have only local network access. Fair-share and allocation policies
are in place both for the PBS batch system and shared file systems.

% Data staging is supported via the DTNs, from the wide area network and from
% the OLCF high performance storage system (HPSS). DTNs are shared across all
% Titan's users that can schedule jobs to a dedicates batch system to automate
% data staging. Fair-share policies are in place both for the batch system and
 %for the use of DTN resources.

Titan's architecture, configuration and policies poses several challenges to the
integration with PanDA. The deployment model of PanDA Pilot is unfeasible on
Titan: PanDA Pilot requires to contact the Job Dispatcher of the PanDA Server to
pull jobs to execute but this is not possible on Titan because worker nodes do
not offer outbound network connectivity. Further, Titan does not support PanDA's
security model based on certificates and virtual organizations, making the
PanDA's approach to identity management also unfeasible. While DTNs offer wide
area network data transfer, an integration with ATLAS DDM is beyond the
functional and administrative scope of the current prototyping phase. Finally,
the specific characteristics of the execution environment, especially the
absence of local storage on the worker nodes and modules tailored to Compute
Node Linux, require reengineering of ATLAS application frameworks.

Currently, no HEP application can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks can be offloaded from
the Grid to Titan's large amount of cores. Further, when HEP tasks can be
partitioned to independent jobs, Titan worker nodes can be used to execute up to
16 concurrent jobs' payload, one for each available core. Given these
constraints and challenges, the type of task most suitable for execution at the
moment on Titan is Monte Carlo
% event generation and
detector simulation.
% Both types of task are
This type of task is mostly computational-intensive, requiring less than 2GB of
RAM at runtime and with small input data requirements. Detector simulation tasks
in ATLAS account for XX--XX\% of all jobs on WLCG, corresponding to X/X of the
current Grid capacity.\mtnote{Please provide numbers to replace `XX' and,
possibly, a diagram illustrating detector simulation stats compared to the whole
ATLAS computation activity.}

Detector simulation is part of the ATLAS production Monte Carlo (MC) workflow
(also known as MC production
chain)~\cite{rimoldi2006atlas,de2013delphes,ritsch2014atlas}. The MC workflow
consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Event generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{sjostrand2006pythia}
and HERWIG~\cite{corcella2001herwig}. The detector simulator is called
Geant4~\cite{agostinelli2003geant4} and simulates the interaction of these
particles with the sensitive material of the ATLAS detector. Each interaction
creates a so-called hit and all hits are collected and passed on for
digitalization where hits are further process to mimic the readout of the
detector. Finally, reconstruction operates local pattern recognition, creating
high-level objects like particles and jets.

@@

The lack of wide area network connectivity on Titan's worker nodes is the most
relevant challenge for integrating PanDA WMS and Titan. Without connectivity,
Panda Pilots cannot be scheduled on worker nodes because they would not be able
to communicate with PanDA Server and therefore pull and execute jobs. This makes
impossible to port PanDA Pilot to Titan while maintaining the defining feature
of the pilot abstraction: decoupling resource acquisition from workload
execution via multi-stage scheduling.

The unavailability of pilots is a potential drawback when executing distributed
workloads like MC detector simulation. Pilots are used to increase the
throughput of distributed workloads: while pilots have to wait in the
supercomputer's queue, once scheduled, they can pull and execute jobs
independently from the system's queue. Jobs can be concurrently executed on
every core available to the pilot, and multiple generations of concurrent
executions can be performed until the pilot's walltime is exhausted. This is
particularly relevant for machines like Titan where queue policies privilege
parallel jobs on the base of the number of worker nodes they request: the higher
the number of nodes, the shorter the amount of queue time (modulo fair-share and
allocation policies).

@@

Titan’s backfill functionality offers the opportunity to avoid the overhead of
queue wait times without using pilot abstraction. Backfill availability is the
number of worker nodes that cannot be used for a certain amount of time by any
of the jobs already queued on Titan: All queued jobs are either too large or
their walltime is too long. At any point in time, Titan’s Moab scheduler can be
queried for backfill availability. Based upon the result of this query, a job
can be shaped to request no more than the backfill availability. As such, when
submitted this job is usually scheduled immediately, spending almost no time in
the queue.

Compared to pilots, backfill has the disadvantage of limiting the amount of work
nodes that can be requested. Pilots are normal jobs: they can request as many
worker nodes for as much time as a queue can offer. On the contrary, jobs sized
on the basis of backfill information availability \jhanote{should be resource
availability or backfill queue information?}\mtnote{AFAIH, there is no dedicated
queue to backfill} depend on the number of worker nodes that cannot be given to
any other job in the Titan's queue.
% \sergeynote{any other job currently in the Titan's queue?}\mtnote{Better?}.

Usually, backfill availability is a fraction of the total capacity of the queue
but the size of Titan mitigates this limitation. Every year, about 10\% of
Titan's capacity remains unused~\cite{titan_utilization}, corresponding to an
average of 30,000 unused cores. This equals approximately 270M core hours per
year, roughly 30\% of the overall capacity of WLCG.

% A system that can use those temporarily free worker nodes with smaller
% scale tasks would be very valuable. For this reason, Titan offers backfill
% functionalities: users can interrogate Titan's Moab scheduler about how many
% free nodes are currently available and for how long. If a user submit a
% requestvfor that number of nodes and walltime, queue time will be minimal.

@@

Given the communication requirements of PanDA Pilots and the unused capacity of
Titan, PanDA pilot was redesigned to serve as a job broker on Titan. Architected
as a stand-alone service to be executed on a DTN, this prototype called `PanDA
Broker' implements functionalities to: (i) interrogate Titan about backfill
availability; (ii) pull MC jobs and events; (iii) wrap the payload of ATLAS jobs
into MPI scripts; (iv) submitting MPI scripts to Titan's PBS batch system and
monitor their execution; and (v) staging input/output files. Backfill querying,
payload wrapping, and scripts submission required a new implementation while
pulling ATLAS job and events, and file staging were mostly inherited from PanDA
Pilot.

% \sergeynote{Monitor job state and state transitions using
% SAGA}\mtnote{Better?}

@@

Backfill querying is performed via a dedicated Moab scheduler command while a
tailored Python MPI script is used to execute the payload of ATLAS jobs. This
MPI script enables the execution of unmodified Grid-centric, ATLAS jobs on
Titan. Typically, a MPI script is workload-specific as it sets up the execution
environment for a specific payload. This involves organization of worker
directories, data management, optional input parameters modification, and
cleanup on exit. Upon submission, a copy of the MPI script runs on every
available worker node,
% Each script knows its MPI rank (an index that runs from zero to a maximum
% number of nodes or script copies) as well as the total number of ranks in the
% current submission. When activated on worker nodes, each copy of the MPI
% script
starting the execution of the ATLAS job's payload in a subprocess and waits
until its completion.

MPI scripts are submitted to Titan's PBS batch system via
RADICAL-SAGA~\cite{radical-saga}, a Python module, compliant with the OGF GFD.90
SAGA specification~\cite{saga-spec}. The Simple API for Grid Applications (SAGA)
offers a unified interface to diverse job schedulers and file transferring
services. In this way, SAGA provides an interoperability layer that lowers the
complexity of using distributed infrastructures. Behind the API façade,
RADICAL-SAGA implements a adaptor architecture: each adaptor interface the SAGA
API with different middleware systems and services, including the PBS batch
scheduler of Titan.

The data staging capabilities of the PanDA Broker are implemented via a file
system that is shared among DTNs and worker nodes. The input files with the
events of the ATLAS jobs are downloaded on the shared filesystem via the ATLAS
DDM service. The MPI script setup process includes making the location of these
files available to the payload of the ATLAS's jobs. The PanDA Broker can locate
the payload's output files on the shared filesystem and transfer them from Titan
to any computing center used by ATLAS.
% the ATLAS Tier 1 computing center at Brookhaven National Lab (BNL) or any
% other Grid site.

Once deployed on Titan, every PanDA Broker supports the execution of MC detector
simulation in 8 steps: (i) PanDA Broker queries Titan's Moab scheduler about
current backfill availability; (ii) the PanDA Broker queries the Job Dispatcher
module of the PanDA server for ATLAS jobs that have been bound to Titan by JEDI;
(iii) the PanDA Broker requests to the Data Service module of the PanDA Server
what input data these ATLAS jobs require and transfer the files to the DTN via
the ATLAS DDM service; (iv) the PanDA Broker creates an MPI script, wrapping
enough ATLAS jobs' payload to fit backfill availability; (v) the PanDA Broker
submits the MPI script to the Titan's PBS batch system via RADICAL-SAGA; (vi)
upon execution on the worker node(s), the MPI script creates a ramdisk and
configure AthenaMP to execute the ATLAS jobs' payload; (vii) the payload is
executed into dedicated processes, 16 for each worker node, 1 for each available
core; (viii) upon completion of the payload execution, the PanDA Broker locates
the output on the shared filesystem and transfer it to designated computing
centre.
